Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene

### 2023-02-05 manuscript/manuscript.Rmd
introduction updates

Existing palaeoclimate reconstructions are generally based on one of three approaches. Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca), and sometimes biological proxies such as leaf physiognomy, are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g. @Liu2009; @Royer2012]. These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...]. Palaeoecological and geological proxy data can provide broad climate constraints, but no established transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Kvacek2007, @Cao2019]. Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

Marine, proxy-based, deep-time palaeoclimate reconstructions commonly rely solely on geochemical proxies [@Veizer2015;@Grossman2022]. Despite the impressive size of these archives large spatial gaps bias global temperature reconstructions based on those records [@Jones2022]. In contrast, studies on the terrestrial and coastal climate record have long used ecological and lithological data for quantitative climate reconstructions [@]. For example, a plethora of "nearest living relative" methods exists for deriving numerical climate estmates from taxonomic plant data for the Cenozoic  [@Faucette2007; @...]. These methods work on the the premise that the climatic requirements of ancient plant taxa were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017].

Even when using the full range of available records, any proxy-based climate reconstructions that goes back beyond the Holocene is inevitably very patchy. To infer regional or global-scale patterns from an incomplete record requires the bridging of spatial gaps. This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled.

Here, we apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.

Combining quantitative proxy data with not fully quantified climate constraints, e.g. the presence of tropical weathering products such as bauxite and laterite, has the potential to improve upon existing, purely proxy-based reconstructions [@Scotese2021; @Burgener2023]. Terrestrial and

@@

To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings. Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g. @Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022]. Reconstructions that take into account the spatial distribution of climate data are... ### HERE###
provide an alternative that can be employed more readily by a broad range of workers. Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].

>>>>

Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g. @Liu2009]. These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...]. 2) Palaeobiological, ecological and geological proxy data can provide broad climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Kvacek2007; @Cao2019; but see @Royer2012]. Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017]. 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g. @Hollis2019], and global temperature reconstructions based on those records alone may be biased [@Jones2022]. To infer regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps. A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019, @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures and precipitation have been created with Gaussian process regression [@Inglis2020; @Chandra2021].

Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. nearest living relative...

We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.

We apply this model to the early Eocene climate optimum,...

"nearest living relative" methods are used for deriving numerical climate estimates from taxonomic plant data for the Cenozoic  [@Faucette2007; @Chevalier2022]. These methods work on the the premise that . 
Proxy-based, deep-time palaeoclimate reconstructions relying solely on geochemical proxies [e.g. @Veizer2015; @Grossman2022] have significant spatial gaps, and global temperature reconstructions based on those records alone may be biased [@Jones2022]. Ecological and lithological data are sometimes used for quantitative climate reconstructions [@Uhl2009;], but are rarely used in conjunction with geochemical proxies [@Burgener2023].

For example, a plethora of

Here, we combine sea surface temperatures from geochemical proxies with numerical temperature estimated from mangroves and coral reefs, derived with a nearest living relative

apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.

, and sometimes biological proxies such as leaf physiognomy,
. This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled. Examples of this approach include t
Combining quantitative proxy data with not fully quantified climate constraints, e.g. the presence of tropical weathering products such as bauxite and laterite, has the potential to improve upon existing, purely proxy-based reconstructions [@Scotese2021; @Burgener2023]. Terrestrial and

@@

To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings. Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g. @Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022]. Reconstructions that take into account the spatial distribution of climate data are... ### HERE###
provide an alternative that can be employed more readily by a broad range of workers.
